The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-22 src/ms.tex
Add back accuracy

Major versions of the core package --- that is, versions that add and/or modify
functionality --- are released approximately every six months, and are then
maintained with releases that fix issues until the next major version is
released. However, every two years a major release is designated as a long-term
support (LTS) release, which continues to be maintained for up to two years
\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
versions of \astropypkg that users who require a high level of stability can
make use of if they do not always need the latest features (this could include,
for example, telescope operation or data reduction pipelines, etc.). The v5.0
release of the core package was designated as LTS, and since it was released at
the end of 2021, it will be maintained until the end of 2023.

@@

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with purposefully few changes to the
software interface. This maturity and stability allows for the broader
astronomy \python community to rely upon the \astropypkg core package and build
specialized packages within the Astropy ecosystem. Even many non-astronomical
\python libraries have come to rely on \astropypkg. The relationship between
\astropy and the \python-using community is reciprocal, with the evolving needs
of the community and the maturation of the broader scientific \python ecosystem
driving much of the development of the \astropypkg core package.

Within \astropypkg, development may be roughly split into a few
categories: new features added, intra- and inter-package interoperability,
improvements to precision, accuracy, and reproducibility, and performance
enhancements. We discuss these in the subsections below.

Additionally, some features have been moved from the core package, either
because they are of more general use outside of \astropypkg, or because they are
too specialized and belong in an Astropy-affiliated package. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
carried, as well as the associated \python wrappers. Since this is basic
infrastructure, with a release cycle set by SOFA, it made more sense to create
separate packages that serve as \astropypkg dependencies: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
was too specialized was a Virtual Observatory sub-package; the Simple
Application Message Protocol was kept in the core package
(\astropysubpkg{samp}), but other parts more properly belonged in
\package{astroquery} (see Section~\ref{sec:astroquery}).

@@

New features present a particular difficulty for a mature package such as
\astropypkg, because it is challenging to know {\em a priori} what the best
interface will be, and hard to address all possible use cases in one round of
development, even within \astropypkg\ itself. New features come in various forms,
from new submodules providing wholly new capabilities, to larger additions to
existing submodules, to large rewrites of the interface of modules. This range
is spanned by five main new features that entered the \astropypkg core since the
previous summary \citepalias{astropy:2018}. In our descriptions of each we
include how we hope the new feature will evolve. Ultimately, in a user-driven
project like \astropy, development depends primarily on the priorities of
contributors.

\paragraph{Uncertainties and Distributions} Early in our careers we were likely
all taught that measurements without units and uncertainties are meaningless.
\astropyunits provides the ability to associate numbers with units, and
propagate these correctly, using the \astropyQuantity\ class. However, the
ability to associate and propagate uncertainties is relatively limited: in
\astropysubpkg{nddata}, there are options to associate errors with data arrays,
but these are treated as independent, i.e., covariances are not tracked.

@@

\astropy now supports reading and writing tables in the American Astronomical
Society Journals' Machine-Readable Table (MRT) format. This ASCII format has
long been missing and as part of this year's Google Summer of Code project
\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}
has been added to the \astropyTable\ I/O. In addition, \astropyTable\ may also read from and write
to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.

The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
standard} has been updated to version 1.0, adding support for three additional data
subtypes. First, it adds multidimensional column data (both masked and unmasked)
with fixed dimensions in all table cells. Second, it adds multidimensional
column data with variable-dimension arrays similar to FITS variable-length
arrays. Third, it adds object-type columns with simple \python objects.

As mentioned previously, \package{dask} arrays may be used as a mixin column in
\astropyTable. Now, \package{dask} may also be the data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and if written to disk, the array
will be computed while written, avoiding excessive memory use. \astropyTable\ can
now be appended to an existing FITS file, and \package{dask} mixin columns
interoperate seamlessly between the two table types.

@@

\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file,
following a given standard. \astropy uses this internally for the I/O methods in
\astropyTable\ and \astropyCosmology, and users can register custom I/O to
extend the options on these classes.

As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
number of new use cases. For instance, the class-based architecture allows for
the creation of custom registries. One application is in \astropycosmology,
which has two different \textit{kinds} of registries for \astropyCosmology: one
for reading and writing files, and another for converting between python
objects. Using \astropyiosubpkg{registry} for custom classes means a developer
does not have to create a whole new I/O system and can offer to the user a
unified and familiar API.

\subsection{Precision, Accuracy, \& Reproducibility}
\label{sec:core-features-precision_accuracy_repreducibility}

>>>>

Major versions of the core package --- that is, versions that add and/or modify
functionality --- are released approximately every six months, and are then
maintained with releases that fix issues until the next major version is
released. However, every two years a major release is designated as a long-term
support (LTS) release, which continues to be maintained for up to two years
\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
versions of \astropypkg that users who require a high level of stability can
make use of if they do not always need the latest features (this includes but
not limited to telescope operations and data reduction pipelines). The v5.0
release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
it will be maintained until the end of 2023.

@@

The \astropypkg core package is mature and stable, in that many features have
been part of the core package for years with purposefully few changes to the
software interface. This maturity and stability allows for the broader
astronomy \python community to rely upon the \astropypkg core package and build
specialized packages within the Astropy ecosystem. Even many non-astronomical
\python libraries have come to rely on \astropypkg. The relationship between
\astropy and the broader \python community is reciprocal, with the evolving needs
of the community and the maturation of the broader scientific \python ecosystem
driving much of the development of the \astropypkg core package.

Within \astropypkg, development may be roughly split into several
categories: new features, intra- and inter-package interoperability,
improvements to numerical precision, accuracy, and reproducibility, performance
and documentation enhancements, and bug fixes. We discuss some of these in the subsections below.

Additionally, some features have been moved from the core package, either
because they are of more general use outside of \astropypkg, or because they are
too specialized and belong in an Astropy-affiliated package. An example of the
former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA) software
Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that \astropy
carried, as well as the associated \python wrappers. Since this is basic
infrastructure, with a release cycle set by SOFA, it made more sense to create
separate packages that serve as \astropypkg dependencies: \package{ERFA}
\citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
was too specialized was a Virtual Observatory sub-package; the Simple
Application Message Protocol (\astropysubpkg{samp}) is kept in the core package
for now, but other parts more properly belonged in
\package{astroquery} (see Section~\ref{sec:astroquery}).

@@

New features present a particular difficulty for a mature package such as
\astropypkg, because it is challenging to know {\em a priori} what the best
interface will be, and hard to address all possible use cases in one round of
development, even within \astropypkg\ itself. New features come in various forms,
from new submodules providing wholly new capabilities, to larger additions to
existing submodules, to large rewrites of the interface of modules. This range
is spanned by five main new features that entered the \astropypkg core since the
previous summary \citepalias{astropy:2018}. In our descriptions of each, we
include how we hope the new feature will evolve. Ultimately, in a user-driven
project like \astropy, development depends primarily on the priorities of
contributors.

\paragraph{Uncertainties and Distributions} In scientific analysis,
measurement values with units and uncertainties are essential.
\astropyunits provides the ability to associate numbers with units, and
propagate these correctly, using the \astropyQuantity\ class. However, the
ability to associate and propagate uncertainties is relatively limited: in
\astropysubpkg{nddata}, there are options to associate errors with data arrays,
but covariances are not tracked.

@@

\astropy now supports reading and writing tables in the American Astronomical
Society Journals' Machine-Readable Table (\href{https://docs.astropy.org/en/stable/io/ascii/write.html#cds-mrt-format}{MRT}) format. This ASCII format has
long been missing and was added to \astropyTable\ I/O in a Google Summer of Code 2021 project.
In addition, \astropyTable\ may also read from and write
to \href{https://asdf-standard.readthedocs.io/en/latest/}{ASDF},
\href{https://docs.astropy.org/en/stable/io/unified.html#table-io-parquet}{Parquet},
and \href{https://wwwastro.msfc.nasa.gov/qdp/}{QDP} formats.

The \href{https://github.com/astropy/astropy-APEs/blob/main/APE6.rst}{ECSV
standard} has been updated to v1.0, adding support of three additional data
subtypes: multidimensional column data (both masked and unmasked)
with fixed dimensions in all table cells, multidimensional
column data with variable-dimension arrays similar to FITS variable-length
arrays, and object-type columns with simple \python objects.

As mentioned previously, \package{dask} arrays may be used as a mixin column in
\astropyTable. Now, \package{dask} may also be a data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
be appended to an existing FITS file, where \package{dask} mixin columns can
interoperate seamlessly between them.

@@

\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file,
following a given standard. \astropypkg uses this internally for the I/O methods in
\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
extend the options on these classes.

As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
number of new use cases; for instance, the class-based architecture allows
streamlining of the creation of custom registries while minimizing code duplication.
One application is in \astropycosmology,
which has two different \textit{kinds} of registries for \astropyCosmology: one
for reading and writing files, and another for converting between \python objects.

\subsection{Numerical Precision, Accuracy, \& Reproducibility}
\label{sec:core-features-precision_accuracy_repreducibility}
